Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis.

### 2023-12-13 paper/paper-new.Rmd
update

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      dev = "pdf",
                      fig.align = "center",
                      message = FALSE,
                      warning = FALSE)
```

```{r packages, include = FALSE}
library(tidyverse)
library(ordinal)
library(here)
library(flextable)
devtools::load_all()
```

@@

@@

Psychological research make an extensive use of ordinal data. One of the main reason is probably the usage of Likert scales [@Likert1932-xw]. Ordinal data refers to a specific type of measurement scale [@Stevens1946-te] where ordered numbers are assigned to a variable. Compared to nominal scale and as the name suggest the labels are ordered. Compared to interval or ratio scales there is no explicit assumption about the distance between labels. (rivedi questo). An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement). Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer. @Stevens1946-te suggest that for ordinal variables is appropriate to calculate ranks-based statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation). This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data. In Psychology especially when using item-based measures (questionnaires, surveys, etc.) the common practice is using a normal linear regression that makes an explicit assumption about metric features of the response variable. @Liddell2018-wu reviewed the psychological literature using likert-based measures and reported how the majority of papers used metric-based statistical models. In the same work, @Liddell2018-wu showed extensive examples and simulations about the potential pitfalls of treating an ordinal variable as metric. They reported problems in terms of lack of power, inversion of the effects and distortion in estimating the effect size (rivedi meglio qui). (vedi se c'è qualche altro lavoro che fa vedere questa cosa).

### Ordinal regression models

Despite the actual modeling proposal by @Liddell2018-wu, there is a class of regression models taking into account the ordinal nature of the response variable without making metric assumptions. We can class this general class of models as *ordinal regression*. The nomenclature of these models can be confusing mainly because there are several subclasses of models with different assumptions and structures [@Tutz2022-dg]. @Tutz2022-dg and @Burkner2019-aw provide a clear and updated taxonomy of ordinal regression models. We can identify three main classes: *cumulative models* [@McCullagh1980-cw], *sequential models* [@Tutz1990-fe] and *adjacent category models*. The cumulative is the mostly used model assuming the existence of a latent variable that when segmented using thresholds produces the observed ordinal variable. The psychological process underlying the response is clearly formalized in the signal detection theory framework where the respondent (vedi come sistemare). The sequential model as suggested by the name is appropriate when modelling sequential processes. Assuming to have 5 response options, the sequential model assume that responding 3 means already reaching the states 1 and 2. A clear example is proposed by @Burkner2019-aw where the marriage duration in years is predicted as a function of some explanatory variables. For each level of the response variable there is a latent distribution where the step between a marriage year $k = 1$ and the next years $k > 1$ is modeled by the sequential regression. When comparing $k$ with $k > 1$, everything lower than $k$ is assumed to be already reached. The adjacent category model compare the category $k$ with $k + 1$ still assuming a latent distribution for each $k$. As suggested by @Tutz2022-dg the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories.

## notation

- idea di mettere $F$ o qualcosa di simile per la pdf ($\Phi$ e.g., per la normale)
- $Y^\star$ per la latente
- $k$ numero di livelli ordinali
- $\alpha$ per le soglie

per notazione https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf e @Burkner2019-aw

per la parte di notazione e quella più formale anche @DeCarlo2010-lj

## cumulative vs others

- cumulative quando una variabile continua latente viene tagliata in modo osservato. Le likert possono essere gestite in questo modo

from @Burkner2019-aw:

> For  many  ordinal  variables,  the  assumption  of  a   single underlying continuous variable, as in cumulative  models, may not be appropriate. If the response can  be understood as being the result of a sequential pro- cess, such that a higher response category is possible  only after all lower categories are achieved, the sequen- tial model proposed by Tutz (1990) is usually appropri- ate.

vedi se fare k-1 logistiche è più o meno la stessa cosa

in realtà credo che l'adiaject category e sequential model comunque abbiamo la proportional odds assumption

```{r}
#| eval: false
library(brms)
library(ordinal)

dat <- data.frame(x = rnorm(100))
dat <- sim_ord_latent(~x, By = log(2), probs = c(0.5, 0.2, 0.2, 0.1), data = dat, link = "logit")

fit_clm <- clm(y ~ x, data = dat, link = "logit")
fit_cum <- brm(y ~ x, data = dat, family = cumulative(link = "logit"))
fit_seq <- brm(y ~ x, data = dat, family = sratio(link = "logit"))
fit_adj <- brm(y ~ x, data = dat, family = acat(link = "logit"))

come sottolinea @Burkner2019-aw e @Tutz2020-xq i sequential model sono adatti per un certo tipo di dati dove l'interesse è vedere l'impatto dei predittori sulla transizione ad un categoria $r$ considerando che la categoria $r - 1$ è stata raggiunta.

@Burkner2019-aw suggest that while cumulative and sequentials models have a clear empirical distinction the adjiacent category model is usually used for computational advantages. The probability of passing from one category to the other is modeled as a function of predictors.

## Risolvere un pochino la nomenclatura

https://cran.r-project.org/web/packages/ordinal/vignettes/clm_article.pdf

@@

Da citare in qualche modo @Kemp2021-dj

> “As a matter of fact, most of the scales used widely and effectively by psychologists are ordinal scales. In the strictest propriety the ordinary statistics involving means and standard deviations ought not to be used with these scales, for these statistics imply a knowledge of something more than the relative rank-order of data. On the other hand for this ‘illegal’ statisticizing there can be invoked a kind of pragmatic sanction: In numerous instances it leads to fruitful results.” (Stevens, 1946, p. 679.)

> It is straightforward to construct ordinal scales that do not involve rank ordering. For example, one can take the first element encountered and arbitrarily assign it the number 100. If the next element encountered is smaller it is given a smaller arbitrary number, 53 say. If the third element is between these two, it can be given the number 86, and so on. If this construction method is used, the difference between the elements assigned the numbers 70 and 80 will not in any important sense be equal to the difference between the elements assigned 90 and 100, and the intervals between the numbers are not really interpretable. Note, too, that monotonic transformations of the scale essentially leave the mea- sure unaffected.

interessante come @Cliff2016-ck descriva che la maggiorparte dellle domande di ricerca in psicologia siano in riferimento alla location di un costrutto e possano essere gestite in modo ordinale

The use of ordinal data is widespread in Psychology. Usually items from questionnaires are created using Likert scales where a certain psychological traits is measured using an intuitive scale with 3-5 or more anchor points. These measure cannot be considered *metric* measure on interval or ratio scales [stevens] but the categories are ordered.

Other examples here

Despite the usage of ordinal variables, statistical models made for these type of data are rarely used in Psychology. @Liddell2018-wu reported that the majority of published papers using likert-like measures used standard methods to analyze the data. In practical terms, they use *metric* models where the response variable cannot be considered fully numeric (vedi se c'è un termine tipo scala a rapporti).

altro sui modelli

Theoretically, using a *metric* for ordinal data is not appropriate but understanding the actual impact is not straightforward. @Liddell2018-wu did a comprehensive work about pitfalls of analzying ordinal data as metric. They showed that metric models produce higher type-1 and type-2 errors compared to the ordinal models. In particular, given the bounded nature of ordinal data, difference in the underlying latent distribution are not always captured by the metric model that simply estimate the mean of the ordinal variable. This is even more relevant when the underluing variance of the ordinal variables are not homogeneous. Furthermore they presented some situations where the metric model could be wrong in the opposite direction, finding an effect with the wrong sign (type-s error).

Finally, they also demonstrated that even in the best condition where e.g. comparing two groups the variance are equal, the ordinal model is more powerful than the metric model given the underestimation of the true effect size from the latter.

> Some authors have argued that, despite the ordinal character of individual Likert items, averaged ordinal items can have an emergent property of an interval scale and so it is appropriate to apply metric methods to the averaged values (e.g., Carifio & Perla, 2007, 2008).

vedi come gestire questo

anche questo è importante
> Ordered-probit models typically assume that the thresholds (θ k) are the same across all groups because the thresholds are theoretically linked to the response measure, not to the predictor value. For example, when asked, “How happy are you?” with response options ‘1’ = very unhappy, ‘2’ = mildly unhappy, ‘3’ = neutral, ‘4’ = mildly happy, ‘5’ = very happy,” the latent thresholds between ordinal levels are assumed to be implicit in the phrasing of the question, regardless of other aspects of the respondent or situation. In other words, the thresholds are assumed to be part of the measurement procedure, not dependent on the value of the predictor or covariate. This can be technically referred to as a type of measurement invariance.

Da citare come paper iniziale sui modelli [@McCullagh1980-cw].

questo per la tassonomia dei diversi modelli ordinali [@Tutz2022-dg]

da vedere magari per una critica tipo Gomila sul binary model [@Robitzsch2020-la]

## Metriv vs ordinal models

- the main difference is that the metric model assign a number to each label of the discrete variable assuming that the distance is the same. [@Liddell2018-wu]

@@

clm(y ~ x1 + x2, data = dat, link = "probit")
coef(fit)/sigma(fit)

@@

## Odds ratios and cumulative odds ratios

- https://online.stat.psu.edu/stat504/lesson/4/4.1

In this section we introduce the main concept to understand the effects of a logit model.

## Demonstrating the proportional odds assumption

basically given that the effect of $\beta$ is constant the odds ratio is independent from the thresholds $\alpha_j$ [@Liu2023-bp]. In the current paper we are simulating data using a single set of $\beta$s thus the proportional odds assumption is true. @Liu2023-bp reviewed the available methods to test this assumption.

- https://hbiostat.org/ordinal/impactpo.pdf blog su proportional odds, mi pare di capire che non sia così problematica come assunzione

sarebbe da capire quali sono i rischi di assumere questo ed eventualmente se simulare non proportional odds è troppo complicato.

Let's simulate the effect of a binary predictor on ordinal scale 1-5:

```{r}
#| echo: true
b1 <- log(3) # log odds ratio
n <- 1e4
x <- rep(c("a", "b"), each = n/2)
dat <- data.frame(x = x)
probs <- rep(1/5, 5) # for the group "a", uniform probabilities
dat <- sim_ord_latent(~x, By = b1, probs = probs, data = dat, link = "logit")
fit <- clm(y ~ x, data = dat, link = "logit")
pr <- predict(fit, data.frame(x = unique(x)))$fit
pr
```
Basically the proportional odds suggest that:

$$
\text log(\frac{P(y \leq 1)}{P(y > 1)})
$$
Is the same regardless the level of the $x$ predictor. Thus:

@@

```{r tab-sim-probs}
truth <- c(alphas, b1)
names(truth) <- c(names(fit$alpha), names(fit$beta))
clm_table(fit, truth = truth, caption = "caption") |> 
  flextable_to_rmd()
```

@@

```{r tab-sim-latent}
truth <- c(alphas, b1)
names(truth) <- c(names(fit$alpha), names(fit$beta))
clm_table(fit, truth)
```

@@

```{r}
#| results: asis

filor::print_fun(funs$sim_ord_latent)
```

@@

```{r}
#| eval: false
truth <- c(alphas, b1, z1)
names(truth) <- c(names(fit$alpha), names(fit$beta), names(fit$zeta))
clm_table(fit, truth)
```

@@

>>>>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      dev = "pdf",
                      fig.align = "center",
                      size = "scriptsize",
                      message = FALSE,
                      warning = FALSE)
```

```{r hooks, include = FALSE}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

```{r packages, include = FALSE}
library(tidyverse)
library(ordinal)
library(here)
library(flextable)
library(cowplot)
library(viridis)
devtools::load_all()
```

@@

```{r ggplot, include = FALSE}
mtheme <- function(){
  ggthemes::theme_pander(15) +
    theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5),
        strip.text = element_text(face = "bold"),
        panel.grid.minor = element_blank())
}

theme_set(mtheme())

# palettes
options(ggplot2.continuous.colour="viridis")
options(ggplot2.continuous.fill = "viridis")
scale_color_discrete <- scale_colour_viridis_d
scale_fill_discrete <- scale_fill_viridis_d
```

```{r include = FALSE}
# figures
source(here("scripts", "figures.R"))
rfigs <- readRDS(here("objects", "r-figs.rds"))
```

```{r tables, include=FALSE}
mtab <- function(data){

}
```

@@

Psychological research make an extensive use of ordinal data. One of the main reason is probably the usage of Likert scales [@Likert1932-xw]. Ordinal data refers to a specific type of measurement scale [@Stevens1946-te] where ordered numbers are assigned to a variable. Compared to nominal scale and as the name suggest the labels are ordered. Compared to interval or ratio scales there is no explicit assumption about the distance between labels. (rivedi questo). An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement). Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer. @Stevens1946-te and @Kemp2021-dj suggested that for ordinal variables is appropriate to calculate ranks-based statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation). This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data. Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels [@Liddell2018-wu].

In Psychology especially when using item-based measures (questionnaires, surveys, etc.) the common practice is using a normal linear regression that makes an explicit assumption about metric features of the response variable. @Liddell2018-wu reviewed the psychological literature using likert-based measures and reported how the majority of papers used metric-based statistical models. In the same work, @Liddell2018-wu showed extensive examples and simulations about the potential pitfalls of treating an ordinal variable as metric [but see @Robitzsch2020-la for an alternative perspective]. They reported problems in terms of lack of power, inversion of the effects and distortion in estimating the effect size (rivedi meglio qui). (vedi se c'è qualche altro lavoro che fa vedere questa cosa). Some authors suggested that despite individual items being ordinal, averaging multiple items could solve the problem of applying metric models (e.g., Carifio & Perla, 2007, 2008 metti queste citazioni). @Liddell2018-wu suggested that in some conditions even averaging multiple items and applying metric models could be problematic. @Cliff2016-ck suggest that most of the research questions in behavioral sciences can be considered as ordinal (*is the score $x$ higher than the score $y$*?) concerning variables where the most appropriate measurement scale is probably ordinal.

qualcosa ancora qui

## Ordinal regression models

Despite the actual modeling proposal by @Liddell2018-wu, there is a class of regression models taking into account the ordinal nature of the response variable without making metric assumptions. We can class this general class of models as *ordinal regression*. The nomenclature of these models can be confusing mainly because there are several subclasses of models with different assumptions and structures [@Tutz2022-dg]. @Tutz2022-dg and @Burkner2019-aw provide a clear and updated taxonomy of ordinal regression models. We can identify three main classes: *cumulative models* [@McCullagh1980-cw], *sequential models* [@Tutz1990-fe] and *adjacent category models*. The cumulative is the mostly used model assuming the existence of a latent variable that when segmented using thresholds produces the observed ordinal variable. The psychological process underlying the response is clearly formalized in the signal detection theory framework where the respondent (vedi come sistemare). The sequential model as suggested by the name is appropriate when modelling sequential processes. Assuming to have 5 response options, the sequential model assume that responding 3 means already reaching the states 1 and 2. A clear example is proposed by @Burkner2019-aw where the marriage duration in years is predicted as a function of some explanatory variables. For each level of the response variable there is a latent distribution where the step between a marriage year $k = 1$ and the next years $k > 1$ is modeled by the sequential regression. When comparing $k$ with $k > 1$, everything lower than $k$ is assumed to be already reached [@Tutz2020-xq]. The adjacent category model compare the category $k$ with $k + 1$ still assuming a latent distribution for each $k$. As suggested by @Tutz2022-dg the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories. @Burkner2019-aw suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical distinction as for the cumulative vs sequential model.

In the current paper we put the focus on the cumulative model for several reasons. The first reason is that the latent formulation of the model is particularly convenient both for parameter interpretation and data simulation. The second reason is that several psychological variables can be formalized as a latent continuous construct that when measured is collected using ordinal items. Furthermore, also the signal detection theory framework models the decision process assuming a latent distribution for the signal and noise where thresholds are the response criteria [e.g., @DeCarlo2010-lj].

```{r fig-ordinal-models, fig.cap=filor::get_caption(), out.width="100%"}
knitr::include_graphics("img/fig-ordinal-models.png")
```

## Model notation

metti qualcosa sulla nomenclatura

@@

In this section we introduce some notations for the cumulative model that is used through the paper and in the R code. The notation is as consistent as possible with the main textbooks and with the `ordinal` R package used in the tutorial. We define $Y_k$ as the observed ordinal response with $1, 2, \dots,k$ options and $Y^\star$ is the corresponding latent variable. The latent variable is segmented using $k - 1$ thresholds $\alpha_1, \alpha_2, \dots, \alpha_{k - 1}$. With $g()$ we define the cumulative probability function assumed for the specific model. For example, when using a Gaussian distribution we are fitting a model with a `probit` link function thus $f() = \Phi()$. With $g^{-1}()$ we define the inverse of the cumulative probability function. The minus sign in $\mathbf{X}\mathbf{\beta}$ is used to give the usual interpretation of standard regression models where an increase in $\beta$ corresponds to an increase in the probability of higher categories of $Y$ (metti il libro di agresti sui dati ordinali).

\begin{equation} 
P(Y \leq k) = g(\alpha_k - \mathbf{X} \mathbf{\beta}), \;\;k = 1, 2, \dots, k
(\#eq:prob-cum-model1)
\end{equation}

We can call the $\mathbf{X}\mathbf{\beta}$ as the linear predictor $\eta$. Equation (num) describe how the probability of a single outcome $Y_k$.

\begin{equation} 
P(Y = k) = g^{-1}(\alpha_k - \eta) -  g^{-1}(\alpha_{k - 1} - \eta)
(\#eq:prob-cum-model2)
\end{equation}

<!-- notazione https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf e @Burkner2019-aw per la parte di notazione e quella più formale anche @DeCarlo2010-lj -->

The same model can be formalized using the latent formulation (see Equation). This is more similar to a standard linear regression.

\begin{equation} 
Y^\star_i = \eta + \epsilon_i
(\#eq:latent-model)
\end{equation}

Where $\epsilon$ is the error part and comes from the distribution that is assumed from the latent model. For a *probit* model, errors are sampled from a standard Gaussian distribution while for a *logit* model from a standard logistic distribution. Then the observed ordinal value $Y_i$ comes from $Y^\star_i$ being between the thresholds $\alpha_{k - 1} < Y^\star_i < \alpha_{k}$ with thresholds organized as $- \infty \equiv \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k \equiv \infty$. <!-- TODO rivedi questa -->

In the basic version of the model, the thresholds $\alpha$ are considered as fixed and do not vary as a function of predictors. The thresholds are part of the measurement procedure [@Liddell2018-wu]. In a more sophisticated version of the model called location-shift [@Tutz2022-dg], both the location $\mu$ and the thresholds $\alpha$ can vary as a function of the predictors. In the following sections another version of the model called location-scale [@Tutz2022-dg; @Rigby2005-ko; @Cox1995-ur] will be presented as a more flexible way to model latent distributions with heterogeneous variance (spiega meglio).

@@

clm(y ~ x1 + x2, data = dat, link = "probit")
coef(fit)/sigma(fit)
```

@@

# Interpreting parameters

## Odds and odds ratio

una piccola spiegazione di odds e odds ratio

## Proportional odds assumption

Following again the taxonomy by @Tutz2022-dg, each of the presented ordinal regression model has a basic version making the  proportional odds assumption. There are more advanced versions of the model relaxing this assumption completely (*non proportional odds* models) and partially (*partial proportional odds* models) (rivedi se la nomenclatura va bene). The proportional odds assumption is a crucial point when intepreting model parameters. Can be formalized as in Equation (). Basically if we use the *logit* link function $g()$ the $\beta$s are interpreted as standard odds ratio in the logistic regression. Given that we have $k > 2$ alternatives there is the need of $k - 1$ equations (as reported in Equation ()). The porportional odds is assuming that the regression coefficients $\beta$ are independent from the thresholds $\alpha$ thus the effect is the same regardless of the $Y$ level. Thus $\beta_1 = \beta_2 \dots = \beta_{k - 1}$. The Figure () depicts the proportional odds assumption for the $k - 1$ logistic curves both for probabilities and linear predictors $\eta$.

```{r fig-prop-odds, fig.cap=filor::get_caption()}
rfigs$fig_prop_odds
```

The proportional odds assumption is convenient because regardless the number of $k$, the $\beta_j$ ($j$ being the number of regression coefficients) effect is assumed to be the same. The model is more parsimonious compared to estimating $k - 1$ coefficients for each $\beta_j$ as in the multinomial regression or the non-proportional version of the ordinal regression. Clearly this can be a strong assumption in some conditions. There are several methods to test the proportional odds assumption [see @Liu2023-bp for an overview]. @Tutz2020-xq suggested that using alternative as the location-shift or location-scale models, the assumption can be made (improving model parsimony) but still having a more flexible modeling framework. In the next section, we provide more details about the location-scale model in terms of parameters interpretation and data simulation.

To make a practical example, the dataset presented in Table contains simulated data from $n = 100$ participants rating the agreement about a certain item with $k = 4$ ordered options. The participants are divided into two groups (predictor $x$). We want to fit a cumulative ordinal model and check the proportional odds assumption.

<!-- - https://hbiostat.org/ordinal/impactpo.pdf blog su proportional odds, mi pare di capire che non sia così problematica come assunzione -->

```{r}
b1 <- log(3) # log odds ratio
k <- 4
n <- 100
x <- rep(c("a", "b"), each = n/2)
dat <- data.frame(x = x)
probs <- rep(1/k, k) # for the group "a", uniform probabilities
dat <- sim_ord_latent(~x, By = b1, probs = probs, data = dat, link = "logit")
fit <- clm(y ~ x, data = dat, link = "logit")
pr <- predict(fit, data.frame(x = unique(x)))$fit
```

Basically the proportional odds suggest that:

$$
\text log(\frac{P(y \leq 1)}{P(y > 1)})
$$

Is the same regardless the level of the $x$ predictor. Thus:

@@

```{r tab-sim-probs}
truth <- c(alphas, b1)
names(truth) <- c(names(fit$alpha), names(fit$beta))
#clm_table(fit, truth = truth)
```

@@

```{r tab-sim-latent}
truth <- c(alphas, b1)
names(truth) <- c(names(fit$alpha), names(fit$beta))
#clm_table(fit, truth)
```

@@

```{r}
#| results: asis
filor::print_fun(funs$sim_ord_latent)
```

@@

```{r}
#| eval: false
truth <- c(alphas, b1, z1)
names(truth) <- c(names(fit$alpha), names(fit$beta), names(fit$zeta))
#clm_table(fit, truth)
```

@@

(ref:fig-ordinal-models) Theoretical figure about ordinal models. Voglio fare una figura come questa in @Burkner2019-aw ma con qualche modifica per fare vedere i tre modelli.

(ref:fig-prop-odds) caption here
